Efficient Simulation of Economic and Ecological Dynamics in Aquatic Systems

### 2022-10-12 marlin-paper.qmd
v0.1

---
title: "Efficient Simulation of Economic and Ecological Dynamics for Marine Spatial Planning"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Lennon Thomas
    - Echelle Burns
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
bibliography: references.bib
params:
  params:
  run_name: ["v1.1"]
---

```{r setup}
#| echo: false
#| message: false

@@

need to contrast with SEAPODYM

@@

Many models have been developed to support the theory and design of
spatial management strategies [@fulton2015]. However, they have tended
to fall into one of either highly complex and tactical models designed
for use in a specific location, or to be highly stylized representations
intended to provide theoretical insights with less relevance for highly
specific design questions. This complex landscape presents a challenge
to stakeholders charged with marine spatial planning exercises lacking
accesible models suited for more tactical applications, and a barrier to
the scientific community, wherein comparisons across spatial management
simulation studies are clouded by discrepancies in underlying model
structures beyond differences in the core phenomenon in question.

In part to help address this challenge, we present here a novel
bio-economic modeling framework called `marlin` that allows for
efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulation the impacts spatial
policies such as MPAs across a range of species targeted by various
fleets across heterogeneous and dynamic seascapes. Here we present the
core methods and functionality of this framework, and demonstrate its
use in two xx case-study applications. In the first, we show how
`marlin` can be used to assess critical drivers for positive outcomes
and unintended consequences of MPAs as a bycatch-mitigation strategy. In
the second, we present a more tactical application, demonstrating how
`marlin` can be used to compare and design MPA networks to achieve
societal objectives.

-   MPAs increasingly part of plan

-   A model based field - hard to do experiments at the time and scale
    needed

-   many models out there, but challenges

    -   highly stylized

    -   highly tactical

-   We present a general tool for modeling spatial dynamics of fished
    populations and fishing fleets, with a focus on MPAs

-   We present two case studies showing the utility of this approach

    -   habitat heterogeneity and bycatch

    -   MPA optimization

-   Basis for improved "what if" testing of MPA scenario

@@

Models of MPAs span a wide range of complexity, from simple analytical
tools to end-to-end ecosystem models [@fulton2015]. `marlin` follows a
"models of intermediate complexity" (MICE) philosophy [@plagányi2014]
targeted specifically around the role of technical interactions across
species. Many MPA models simulate the impacts of MPAs on a
single-species level [@ovando2021a]. This can be useful for for example
examining the ways in which species-specific life history traits
interact with fleet dynamics to influence MPA outcomes. However, bycatch
problems are driven by technical interactions, in which fishing effort
for a target species incidentally captures another species. `marlin`
accounts for this by allowing the user to simulate an arbitrary number
of fleets capable of catching an arbitrary number of species in a
spatial area. Different fleets can place different economic value on
different species, while also having species-specific selectivity
functions. This allows us to simulate bycatch dynamics of MPAs in
pelagic ecosystems.

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses.

@@

The key feature of `marlin` for the purposes of this project is the
ability to capture technical interactions across species. While marlin
does not simulate species interactions, fishing fleets in `marlin` are
capable of targeting and affecting multiple species simultaneously. This
allows us to capture for example the potential impacts of displacement
of effort from one area into another across multiple species.

A fishing fleet or *metier* is defined in `marlin` by a total pool of
effort that can have different catchability, prices, and selectivity for
individual species within a simulation. For example, both a longline and
purse seine fleet may capture bigeye (*Thunnus obesus*) and skipjack
(*Katswamus pelamus*), but the longline fleet is much more likely to
catch bigeye than skipjack, and *vice versa*. Each fishing fleet
generates catches, revenues, and profits from individual species that it
targets, and makes decisions based on its total objective function
across all species.

@@

$$
 R_{t,p,f} = \sum_s^S \sum_a^A \Pi_{s,f}\frac{\alpha_{a,s,f}q_{s,f}E_{t,p,f}}{F_{t,p,s,a}} \times Y_{t,p,s,a}
$$

where *p* is patch, *s* is species, *a* is age, $\alpha$ is selectivity
at age, *q* is catchability , *F* is total fishing mortality, $\Pi$ is
price per unit weight, and *Y* is total catches.

$$
Y_{t,p,s,a} = \frac{F_{t,p,s,a}}{z_{t,p,s,a}} \times b_{t,p,s,a} \times (1 - e^{-z_{t,p,s,a}}) 
$$

where *F* is total fishing mortality, given by

$$
F_{t,p,s,a} = \sum_f^F\alpha_{a,s,f}q_{s,f}E_{t,p,f}
$$

@@

```{=tex}
\begin{equation}
\alpha_{a,s,f}=\frac{1}{(1 + e^{-log(19)\times\frac{l_{a,s} - l^{sel}_{s,f}}{\delta^{sel}_{s,f}}})}
  (\#eq:sel)
\end{equation}
```
where $l$ is length at age, $l^{sel}$ is the length at 50% selectivity,
and $\delta$ is the difference between the length at 50% selectivity and
the length at 95% selectivity.

@@

$$
 z_{t,p,s,a} = F_{t,p,s,a} + m_{s,a}
$$

@@

This means that each species experiences a total mortality at age in a
time step in a patch, individual fractions of which are portioned off as
catches and subsequently revenues for each fleet. Given revenues, we
then calculate profits $\phi$ per

$$
\phi_{t,p,f} = R_{t,p,f} - c_f(E_{t,p,f} + 1)_{t,p,f}
^{1.3}
$$

where *c* is cost per unit effort and, the 1.3 exponent is in the manner
of @costello2016a . The addition of one to the total effort is to
constrain the value to be greater than 1, preventing undesired behavior
where fishing effort actually gets less expensive if total effort is
less than 1.

The important feature here then is that the revenues and profits for a
given fleet are a function of the prices and catchability of individual
species, but that mortality itself is independent of price. So, a fleet
that has selectivity and catchability greater than zero can exert
fishing mortality on a species even if it obtains no value from that
species.

Each fishing fleet then decides how to allocate its effort in space
based on one of four possible spatial allocation strategies. The ideal
free distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However, the
IFD presents a number of complications for our study. First, in order to
simplify the assessment of the impacts of different fishing mortality
rates we simulate scenarios with a fixed total pool of effort. As MPAs
change the spatial distribution of profits though, the total amount of
effort required to satisfy the IFD may not equal the total amount of
effort required by the model. Second, even if we were to implement the
IFD for a single fleet, the model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would be a
game-theoretic outcome of the actions of each of the individual fleet
conditional on the actions of all other fleets. While possible to
implement this would slow down our model runs to the point of
practically preventing large-scale evaluation of the effects of MPAs on
bycatch.

@@

```{=tex}
\begin{equation}
n_{t,p,s,a}=\begin{cases}
      = BH(ssb_{t-1,p,s,a}) & \text{if $a = 1$}\\
     = n_{t-1,p,s,a-1}e^{-(z_{t-1,p,s,a-1}}, & \text{if $1< a < max(age)$}\\
     =  n_{t-1,p,s,a}e^{-(z_{t-1,p,s,a})} + n_{t-1,a-1}e^{-(z_{t-1,p,s,a-1})}, & \text{if $a = max(a)$}
  \end{cases}
  (\#eq:pop)
\end{equation}
```
where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass.

@@

```{=tex}
\begin{equation}
  l_{a,s} = l\infty_s\left(1 - e^{-k_s(a - a0_{s})}\right)
  (\#eq:len)
\end{equation}
```
Biomass *b* at age is then given by

```{=tex}
\begin{equation}
  b_{a,s} = wa_s \times l_{a,s}^{wb_s}
  (\#eq:weight)
\end{equation}
```
and maturity *mat* is calculated as

```{=tex}
\begin{equation}
  mat_{a,s} = \frac{1}{(1 + e^{-log(19)\times\frac{l_{a,s} - lmat_s}{{\delta}mat_s}}}
  (\#eq:mat)
\end{equation}
```
where $l_{mat}$ is the length at which on average 50% of individuals are
sexual maturity, and $\delta_{mat}$ is the units of length beyond
$l_{mat}$ at which on average 95% of fish are sexually mature.

@@

```{=tex}
\begin{equation}
  ssb_{t,p,s} = \sum_{a=1}^Aw_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
    (\#eq:ssb)
\end{equation}
```
#### Recruitment {.unnumbered}

@@

```{=tex}
\begin{equation}
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p}}\times{h}\times\sum_{p=1}^P{ssb_{t-1,p}}}{0.2\times{\sum_{p=1}^Pssb0_p}\times(1 - h)+(h - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p}}}\right)\times {hab_p}
  (\#eq:dd2)
\end{equation}
```
where *hab* is a vector of recruitment habitat quality by patch degined
by

$$hab_p = r0_p / \sum_p^P{r0_p}$$ 2. Local density dependence: Density
dependence occurs independently in each patch and recruits are retained
in their home patch.

```{=tex}
\begin{equation}
  n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)
  (\#eq:dd1)
\end{equation}
```
3.  Local density dependence then disperse: Density dependence occurs
    independently in each patch and recruits are then dispersed.

```{=tex}
\begin{equation}
  n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)\times \boldsymbol{d^l} \times \epsilon_t
  (\#eq:dd3)
\end{equation}
```
where **d^l^** is the recruitment movement matrix

@@

```{=tex}
\begin{equation}
  larv_{t,p} = ssb_{t-1} \times\boldsymbol{d^l}
  (\#eq:larvmove)
\end{equation}
```
```{=tex}
\begin{equation}
n_{t,a = 1,p} = \left(\frac{0.8{\times}r0_{p}\times{h}\times{larv_{t,p}}}{0.2\times{ssb0_p}\times(1 - h)+(h - 0.2)\times{larv_{t,p}}}\right) \times \epsilon_t
  (\#eq:dd4)
\end{equation}
```
#### Dispersal

`marlin`'s movement dynamics are based on a continuous-time Markov
chain, as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat).

xx movement equations xx

While this approach to movement simulation is not commonly seen in the
marine modeling literature, it has numerous advantages that warrant its
seeming complexity. First, when only diffusion is present, the model
will generalize to the familiar dynamics of a Gaussian dispersal kernel
xx check this xx. However, the time component *t* in equation xx allows
the movement model to be run on arbitrary time steps while preserving
its behavior.

Second, the *advection* component allows users to input useful features
such as the outputs of ROMs models.xxx

Third, the *taxis* model allow for more active habitat choices by
species, allowing us to simulate both preferences of species in space
and time efficiently.

In this way then, `marlin` allows users to run anything from a simple
Gaussian dispersal kernel up to a system governed by species that
passively diffuse around a dynamic thermal range

Lastly, this form has the advantage that the parameters are directly
estimable from real data. So, if provided with say CPUE data and a
tagging study, users can estimates the movement parameters in the same
manner as @thorson2021, and then pass their estimated parameters to
`marlin` for simulation.

### Rule-of-thumb MPA Design Strategies

MPA design is governed by two general parameters: the number of cells to
place inside an MPAs, and the design strategy used to place the MPAs.

We consider `r n_distinct(experiment_obj$placement_strategy)` different
MPA placement strategies.

1.  Depletion: MPA locations are prioritized in proportion to the
    spawning stock biomass of the most depleted species pre-MPA. So,
    patches with high spawning biomass of highly depleted species are
    prioritized over patches with high spawning biomass of lightly
    depleted species.

2.  Rate: MPAs are placed in based on the pre-MPA depletion-weighted
    catch relative to the total catch in a patch. So, patches with high
    rates of catch of depleted species relative to total catch are
    prioritized.

3.  Avoid_fishing: MPAs locations are prioritized inversely to fishery
    catches. So, patches with low catches are prioritized over patches
    with high catches.

4.  Target_fishing: MPAs locations are prioritized proportionally to
    fishery catches. So, patches with high total catches are prioritized
    over patches with low catches.

5.  Area: For the "area" strategy, we simply assign patches in
    sequential order until the desired total MPA size is reached.

<!-- The information used to inform each MPA design strategy (except area) -->

<!-- can include a multiplicative observation error term $\sigma_{obs}$ where -->

<!-- $$information \sim lnorm(data,\sigma_{obs}) $$ -->

<!-- For example, for the depletion scenario, $\sigma_{obs}$ controls the -->

<!-- accuracy of the spatial depletion data used to prioritize MPA placement. -->

## Reference Points

A note here WHy no reference points? The problem: Fmsy and MSY are
selectivity dependent, and once we allow for different fleets with
different selectivities that critical can move in space, that movement
affects the net selectivity and fishing mortality on the population.

@@

SSBmsy is slightly more conserved, as over the long term the key things
is the flow of recruits, so we do provide an option for that

There are many different strategies for estimating reference points
conditional on selectivity and regimes, see references. We leave it to
users to find relevant reference point as required by their methods on
their own, using the sort of MSE examples available in the package.

## MPA Optimization

@@

$$obj_p = \Phi \times scale(\sum_f^F \boldsymbol{\phi_{f,p}}) + (1 - \Phi)  scale(\boldsymbol{\sum_s^S\frac{ssb_s}{ssb0_{s,p}}})$$

@@

## Generalized Experiments

Two features of marine ecosystems that are often omitted from
theoretical and simulation-based considerations of marine spatial
planning are *technical interactions* and *species on the move*. May
canonical MPA models, informed by planning efforts centered around
relatively static features such as coral reefs, have assumed that
species inhabit a fixed and often uniform (e.g. @hastings2017). In
reality, the habitat of many species are heterogeneous in space, and can
experience both seasonal and directional changes, e.g. climate-change
mediated range shifts.

This reality creates a basic challenge to MPA design, in that given the
need to protect an area less than the total range of a species, the more
complex and dynamics the habitat of the species the more challenging it
will be to identify an area that provides effective protection today and
tomorrow. However, *technical interactions* interact with this spatial
complexity to create another and potentially more difficult challenge.
One of the key questions around spatial management strategies such as
MPAs is to what extent do they *remove* vs *displace* threats.

The potential benefits of MPAs as a bycatch management are complicated
by two key issues: effort displacement and technical interactions. One
of the primary questions around MPAs is to what extent do MPAs remove
versus displace fishing effort. If MPAs remove fishing effort, then
their direct effects on target and bycatch species can only be positive.
If however MPAs displace fishing effort, evaluating their potential
impacts becomes more complicated. The net conservation impact of MPAs
becomes a balancing act between the effects of reduced fishing inside
protected borders and increased fishing intensity outside MPAs. The
outcomes of this displacement can be further complicated by technical
interactions, i.e. the degree to which fishing effort for one species
affects another, particularly when species have markedly different
habitat distributions. For example, @abbott2012 demonstrated that an MPA
designed to protect red king crab (*Paralithodes camtschaticus*) caused
an unintended increase in bycatch of Pacific halibut (*Hippoglossus
stenolepis*) by the displaced bottom-trawl fishery.

Given the complexity of the operating model used here, a complete
sensitivity analysis across all parameters is not practical. Instead, we
focus on the role of life history, fishing pressure, and MPA design.

>>>>

---
title: "Efficient Simulation of Economic and Ecological Dynamics in Aquatic Systems"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Echelle Burns
    - Lennon Thomas
    - James Thorson
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    fig-pos: 'H'
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
bibliography: references.bib
params:
  run_name: ["v1.2"]
abstract: "Very rough cut. Communities around the world are increasingly looking to spatial strategies for managing marine ecosystems. However, the design of spatial management strategies faces numerous challenges. Marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond to changes in marine ecosystems in highly complex ways. We present a framework for efficient simulation of the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate the model's capabilities by focusing on three case studies on the conservation and food security impacts of marine protected areas (MPAs)."
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers

---

```{r setup}
#| echo: false
#| message: false
#| include: false

@@

<!-- Single column, double line spaced -->

<!-- Within the word count (no more than 7000-8000 words for Standard Articles, 3000-4000 words for Applications and Practical Tools) -->

<!-- Continuous line and page numbering throughout -->

<!-- Clearly defined manuscript structure as standard: Author details, Abstract (must be numbered 1-4 according to Manuscript Specifications), Keywords, Introduction, Materials and Methods, Results, Discussion, Figures and Tables with captions -->

<!-- Figures and Tables can be embedded within the text where referenced to facilitate reviewing -->

<!--         Author Contributions (if >1 author – see Authorship Policy). -->

<!--         Statement of where you intend to archive your data -->

@@

Many models have been developed to support the theory and design of
spatial management strategies (see @fulton2015 for a review of existing methods). However, they have tended
to fall into one of either highly complex and tactical models designed
for use in a specific location, or to be highly stylized representations
intended to provide theoretical insights with less relevance for highly
specific design questions. This complex landscape presents a challenge
to stakeholders charged with marine spatial planning exercises lacking
accessible models suited for more tactical applications, and a barrier
to the scientific community, wherein comparisons across spatial
management simulation studies are clouded by discrepancies in underlying
model structures beyond differences in the core phenomenon in question.

In part to help address this challenge, we present here a novel
bio-economic modeling framework called `marlin` that allows for
efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulation the impacts spatial
policies such as MPAs across a range of species targeted by various
fleets across heterogeneous and dynamic seascapes. Here we present the
core methods and functionality of this framework, and demonstrate its
use in three xx case-study applications. In the first, we show how
`marlin` can be used to assess critical drivers for positive outcomes
and unintended consequences of MPAs as a bycatch-mitigation strategy. In
the second, we present a more tactical application, demonstrating how
`marlin` can be used to compare and design MPA networks to achieve
societal objectives.

@@

`marlin` follows a "models of intermediate complexity" (MICE) philosophy [@plagányi2014]
targeted specifically towards capturing heterogeneous habitats, movement dynamics, and technical interactions across species. In its general form, `marlin` simulates the behavior of populations of spatially explicit biologically independent age-structured animal populations affected by one or more fishing fleets in time steps specified by the user (generally annual or finer).  While `marlin`
does not simulate species interactions, fishing fleets in `marlin` are
capable of targeting and affecting multiple species simultaneously. This allows the model to simulate processes such as fisheries bycatch or effort displacement in a way that accounts for the ways that changes in fleet behavior might affect multiple species in the system. The ability to track the impacts of fishing fleets across multiple species simultaneously is particularly important as very few fishing fleets are truly single species.

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note that `marlin` is a complex package with many options to control detailed behaviors. Given that these specific features will evolve over time, we do not describe in detail every option of the model here. Rather, we focus on the equations and methods describing the core foundations of our framework for efficiently simulating bio-economic dynamics in aquatic systems.

@@

A fishing fleet is defined in `marlin` by a total pool of
effort that can have different catchability, prices, and selectivity for
individual species within a simulation (each of which we denote as a *metier*. For example, both a longline and
purse seine fleet may capture bigeye (*Thunnus obesus*) and skipjack
(*Katswamus pelamus*) tunas, but the longline fleet is much more likely to
catch bigeye than skipjack, and *vice versa*. Each fishing fleet
generates catches, revenues, and profits from individual species that it
targets, and makes decisions based on its total objective function
across all species.

@@

$$
 R_{t,p,f} = \sum_s^S \sum_a^A \Pi_{s,f}\frac{\alpha_{a,s,f}q_{s,f}E_{t,p,f}}{u_{t,p,s,a}} \times Y_{t,p,s,a}
$$ {#eq-revenue}

where *p* is patch, *s* is species, *a* is age, $\alpha$ is selectivity
at age, *q* is catchability , *u* is total fishing mortality, $\Pi$ is
price per unit weight, and *Y* is total catches.

$$
Y_{t,p,s,a} = \frac{u_{t,p,s,a}}{z_{t,p,s,a}} \times b_{t,p,s,a} \times (1 - e^{-z_{t,p,s,a}})
$$ {#eq-yield}

where *b* is biomass and *u* is total fishing mortality, given by

$$
u_{t,p,s,a} = \sum_{f=1}^F\alpha_{a,s,f}q_{s,f}E_{t,p,f}
$$ {#eq-u}

@@

$$
\alpha_{a,s,f}=\frac{1}{(1 + e^{-log(19)\times\frac{l_{a,s} - l^{sel}_{s,f}}{\delta^{sel}_{s,f}}})}
$$ {#eq-selectivity}

where $l$ is length at age, $l^{sel}$ is the length at 50% selectivity,
and $\delta$ is the difference between the length at 50% selectivity and
the length at 95% selectivity.

or through a dome shaped form of

XX add dome shaped equation XX

@@

$$
 z_{t,p,s,a} = u_{t,p,s,a} + m_{s,a}
$$ {#eq-z}

@@

This means that each species experiences a total mortality at age in a
time step in a patch, individual fractions of which are portioned off as
catches and subsequently revenues for each fleet. Given revenues *R*, we
then calculate profits $\phi$ per

$$
\phi_{t,p,f} = R_{t,p,f} - C_{t,p,f}
$$ {#eq-profits}

where costs *C* are calculated

$$C_{t,p,f} = \gamma_f  \left( E_{t,p,f}^{\beta_f} + \eta_{f,p} E_{t,p,f} \right)$$

where $$\gamma_f$$ the scales the cost per unit effort for fleet *f*, $\beta$ allows for the cost of effort to scale non-linearly, and $\eta$ is an additional cost per unit effort of fishing in each patch *p*, allowing for the model to account for travel costs for different patches. Travel costs are calculated based on the euclidean distance of each patch to the nearest port of a given fleet, users can specify any number from zero to the number of patches of port locations. When no ports are specified, travel costs are zero.

#### Calculating Total Effort

`marlin` allows for two general modes of effort. The simplest is "constant effort", in which the total effort of each fishing fleet remains constant over time.

The more complex option is "open access". Under open access, the total amount of effort in a given time step for fleet *f* is a function of the profitability of that fleet in the proceeding time steps in which fishing was open.

$$
E_{t+1,f} = E_{t,f} \times e^{\theta_f log(R_{t,f} / C_{t,f})} 
$$ {#eq-oa}

There $\theta$ controls the responsiveness of fleet *f* to the log of the ratio of total revenues *R* to total costs *C*, and approximates the percentage amount effort changes in response to a one unit changes in the log revenue to cost ratio.

Three general exceptions exist for the effort dynamics. When MPAs are implemented, effort that used to operate inside the MPAs can either leave the fishery, or be redistributed outside the the MPA (the default behavior). The model can also accept total catch quotas for each species in each time step. When quotas are activated, if the total catch across all fleets for a given species under the model's effort dynamics would exceed the allowable quota for that species, the total effort for each fleet is decrease proportional to its contribution to the total catch until the quota is satisfied. `marlin` also allows users to specify closed fishing seasons for one or more species in the system. During closed seasons, no effort is allocated to patches that would result in capture of the covered species, potentially resulting in zero effort if the species cannot be avoided. Under these conditions, the amount of effort in the next open season is a function of the effort and profits in the last open fishing season.

#### Tuning Fishing Fleets

The degree of fishing pressure exerted by a given set of fishing fleets on each species is a function of a range of parameters including the total amount of effort, the selectivity ogives, cost functions, the spatial distribution of the resource, the relative prices across species, and the catchability coefficient of each fleet for each species. `marlin` provides a tuning option to help users achieve desired biological outcomes from their fleets.

Users can tune their fleets in one of two ways. First, they can specify a target exploitation rate *u* for each species in their model run. Taking all the other parameters of the model as given, `marlin` then adjusts the catchability coefficients *q_{f,s}* for each fleet *f* and species *s* such that the equilibrium exploitation rate for each species matches the desired level. Second, they can specify a target depletion, defined as total spawning stock biomass under fishing divided by total unfished spawning stock biomass, and the model will adjust the catchability coefficients *q_{f,s}* for each fleet *f* and species *s* such that the equilibrium depletion for each species matches the desired level.

#### Spatial Allocation of Effort

Each fishing fleet decides how to allocate its effort in space
based on one of four possible spatial allocation strategies. The ideal
free distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However, the
IFD presents a number of complications for our model. First, under the constant effort model, the total amount of effort required to satisfy the IFD may not equal the total amount of
effort required by the model. Second, even if we were to implement the
IFD for a single fleet, the model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would be a
game-theoretic outcome of the actions of each of the individual fleet
conditional on the actions of all other fleets. While possible to
implement this would slow down our model runs to the point of
practically preventing large-scale evaluation of the effects of MPAs on
bycatch.

@@

$$
n_{t,p,s,a} = \begin{cases}
      = BH(ssb_{t-1,p,s,a}) & \text{if $a = 1$}\\
     = n_{t-1,p,s,a-1}e^{-(z_{t-1,p,s,a-1}}, & \text{if $1< a < max(age)$}\\
     =  n_{t-1,p,s,a}e^{-(z_{t-1,p,s,a})} + n_{t-1,a-1}e^{-(z_{t-1,p,s,a-1})}, & \text{if $a = max(a)$}
  \end{cases}
$$ {#eq-pop}

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass.

@@

$$
  l_{a,s} = l_{\infty,s}\left(1 - e^{-k_s(a - a0_{s})}\right)
$$ {#eq-len}

Biomass *b* at age is then given by

$$
  b_{a,s} = wa_s \times l_{a,s}^{wb_s}
$$ {#eq-weight}

and maturity *mat* is calculated as

$$
  mat_{a,s} = \frac{1}{(1 + e^{-log(19)\times\frac{l_{a,s} - lmat_s}{{\delta}mat_s}}}
$$ {#eq-mat)}

where $l_{mat}$ is the length at which on average 50% of individuals are
sexual maturity, and $\delta_{mat}$ is the units of length beyond
$l_{mat}$ at which on average 95% of fish are sexually mature.

@@

$$
  ssb_{t,p,s} = \sum_{a=1}^Aw_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
$$ {#eq-ssb}

#### Recruitment

@@

$$
n_{t,p,s,a = 1} = \left(\frac{0.8{\times}\sum_{p=1}^P{r0_{p}}\times{h}\times\sum_{p=1}^P{ssb_{t-1,p}}}{0.2\times{\sum_{p=1}^Pssb0_p}\times(1 - h)+(h - 0.2)\times{\sum_{p=1}^Pssb_{t-1,p}}}\right)\times {hab_p}
$$ {#eq-dd2}

where *hab* is a vector of recruitment habitat quality by patch degined
by

$$hab_p = r0_p / \sum_p^P{r0_p}$$

2. Local density dependence: Density
dependence occurs independently in each patch and recruits are retained
in their home patch.

$$
n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)
$$ {#eq-dd1}

3.  Local density dependence then disperse: Density dependence occurs
    independently in each patch and recruits are then dispersed.

$$
  n_{t,p,s,a = 1,p} = \left(\frac{0.8{\times}r0_{s,p}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p}}}\right)\times \boldsymbol{d^l} \times \epsilon_t
$$ {#eq-dd3}

where **d^l^** is the recruitment movement matrix

@@

$$
  larv_{t,p} = ssb_{t-1} \times\boldsymbol{d^l}
$$ {#eq-larvmove}

$$
n_{t,a = 1,p} = \left(\frac{0.8{\times}r0_{p}\times{h}\times{larv_{t,p}}}{0.2\times{ssb0_p}\times(1 - h)+(h - 0.2)\times{larv_{t,p}}}\right) \times \epsilon_t
$$ {#eq-dd4}

#### Dispersal

`marlin`'s movement dynamics are based on a continuous-time Markov
chain (CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the diffusion and taxis components of this model, though future extensions could easily incorporate advection vectors from say ROMs models (citations). In this way then, `marlin` allows users to run anything from a simple Gaussian dispersal kernel up to a system governed by species that
passively diffuse around a dynamic thermal range.

@thorson2021a described the exact steps of the CTMC approach in detail. We briefly the general CTMC method here, but see @thorson2021a for a detailed description of the method. Under this framework, movement from each patch to each other patch in the system in a given timestep *t* for life stage *g* of species *s* is defined by a movement matrix $\boldsymbol{M}_{t,s,g}$. $\boldsymbol{M}_{t,s,g}$ is calculated as the matrix exponent of the sum of the diffusion and taxis matrices scaled by the width of the time step that the model is running under.

$$
\boldsymbol{M}_{t,s,g} = e^{\Delta t(\boldsymbol{D}_{t,s,g} + \boldsymbol{X}_{t,s,g})} 
$$ {#eq-movement}

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is the taxis matrix.

The diffusion matrix is a function of a user specified diffusion rate, see @thorson2021a. We slightly modify the taxis matrix in the following manner

xx need to fix notation somewhere here, see this line of code for much clearer depiction of what's happening, since sometimes code is cleaner. https://github.com/DanOvando/marlin/blob/3563e7319ad0267f89789267a4c6ee58a9ad0d60/R/fish_class.R#L525 xx

$$
n_{t,p,s,a} = \begin{cases}
      = (1 + \lambda d_{s,f})max(0, h(p2_{t,s,g}) - h(p1_{t,s,g})) & \text{if p2 and p1 are adjacent}\\
     = -\sum_{g' \neq g1}\{ (1 + \lambda d_{s,f})max(0, h(p2_{t,s,g}) - h(p1_{t,s,g}))\} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

The $lambda d_{s,f}$ term is a scalar in which the magnitude of the habitat gradient between patches can be scaled relative to the diffusion parameter $d$.

For both the diffusion and taxis matrix, we allow for the inclusion of physical barriers to movement (i.e. land). Pairs of patches that are adjacent but in which one or both patches are a barrier to movement are set as non-adjacent. The CTMC model then produces movement dynamics that move around barriers rather than over them. This is different than setting habitat in a patch to zero, which implies a patch that a given animal will not move towards for its own sake, but can move over towards another better patch.

While this CTMC approach to movement simulation is to date not commonly seen in the
marine modeling literature, it has numerous advantages that warrant its
seeming complexity. First, when only diffusion is present, the model
will generalize to the familiar dynamics of a Gaussian dispersal kernel. However, the time component *t* in @eq-movement allows the movement model to be run on arbitrarily fine time steps. Second, the *taxis* model allow for more active habitat choices by
species, allowing us to simulate both preferences of species in space
and time efficiently.Lastly, this form has the advantage that the parameters are directly
estimable from real data. So, if provided with say CPUE data and a
tagging study, users can estimates the movement parameters in the same
manner as @thorson2021a, and then pass their estimated parameters to
`marlin` for simulation.

#### Reference Points

Sigh. I need to work on this section, but I am *le tired*.

A note here Why no reference points? The problem: Fmsy and MSY are
selectivity dependent, and once we allow for different fleets with
different selectivities that critical can move in space, that movement
affects the net selectivity and fishing mortality on the population.

@@

SSBmsy is more conserved, as over the long term the key things
is the flow of recruits, so we do provide an option for that, but I haven't messed with it for a while.

There are many different strategies for estimating reference points
conditional on selectivity and regimes, see citations. We leave it to
users to find relevant reference point as required by their methods on
their own, using the sort of MSE examples available in the package.

### MPA Optimization

@@

$$
obj_p = \Phi \times scale(\sum_f^F \boldsymbol{\phi_{f,p}}) + (1 - \Phi)  scale(\boldsymbol{\sum_s^S\frac{ssb_s}{ssb0_{s,p}}})
$$ {#eq-obj}

@@

### Use Cases

We include three examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of marine
protected areas design in a heavily fished coral reef ecosystem. In the
second, we demonstrate how `marlin` can be used to assess components of
climate resilience of alternative MPA design strategies. In the third,
we show how `marlin` can be used to run large-scale simulation testing
of management strategies.

#### Rule-of-thumb MPA Design Strategies

Across all of our case studies we make use of four potential "rule of
thumb" MPA design strategies:

1.  Rate: MPAs are placed in based on the pre-MPA depletion-weighted
    catch relative to the total catch in a patch. So, patches with high
    rates of catch of depleted species relative to total catch are
    prioritized.

2.  Avoid_fishing: MPAs locations are prioritized inversely to fishery
    catches. So, patches with low catches are prioritized over patches
    with high catches.

3.  Target_fishing: MPAs locations are prioritized proportionally to
    fishery catches. So, patches with high total catches are prioritized
    over patches with low catches.

4.  Area: For the "area" strategy, we simply assign patches in
    sequential order until the desired total MPA size is reached.

MPAs never to our knowledge randomly placed; their locations are decided
based on stakeholder negotiation based on available information and
objectives. In theory the design MPA networks can be optimized through
the use of a modeling framework, and indeed conditional on the validity
of the model itself this process is likely to produce better outcomes
than manually-designed strategies [@rassweiler2012; @rassweiler2014].
However, in practice the ability of communities to conduct numerical
optimization of MPA networks can be constrained by the number of
objectives and dimensionality of the operating model, or by simple
capacity of local stakeholders. As such, we show here how `marlin` can
be used to examine the performance of potentially sub-optimal but more
broadly usable design strategies xx.

#### Coral Reefs

In our coral reef example, we model the dynamics of four species: a
grouper (XX), a shallow-reef snapper (xx), a deep-water snapper (XX),
and a reef shark (XX). Groupers undergo a mass migration to a spawning
aggregation once per year, followed by the sharks. Shallow-reef snappers
stay in reefs closer to shower year-round, while deep-water snappers
stay in the deeper reefs (@fig-coral).
